---
title: "MGK OS WS 3 - Rmd practice"
author: "Christoph Scheffel"
date: 'July, 2022'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Rmarkdown practice

In the following part, you will work on three tasks to gain a first insight into formatting text with Rmarkdown.


```{r variables}
set.seed(123)             # set custom seed for reproducible results
x = rnorm(100)            # generate random variable x with 100 obs.
y = 0.2 * x + rnorm(100)  # generate second random variable y with 100 obs. that covaries with x
z = 0.3 * x + rnorm(100)  # generate third random variable z with 100 obs. that also covaries with x
```


Tasks:

1. Format all statistical characters and equations correctly in the following text.

2. Calculate a regression model in the r chunk "regression" using the generated random variables x, y, and z.

3. Write a short paragraph on the results of the linear regression model and report all necessary statistical values.


In the r chunk "variables" in this document, three random variables were generated. 
Each variable has 100 observations.
The variables were created so that the variables y and z covary with x.
In the next r chunk, chunk "regression", a simple linear regression model is calculated.
The formula describing the model is x ~ y + z.


```{r regression}
# compute regression model
model <- lm(x ~ y + z)
summary(model)
```

