---
title: "MGK OS WS 3 - Rmd practice"
author: "Alexander Strobel and Christoph Scheffel"
date: 'July, 2022'
output: pdf_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Rmarkdown practice

In the following part, you will work on four short tasks to gain a first insight into formatting text with Rmarkdown.
Your first task is to format the text in this document so that all statistical characters and equations are displayed correctly.
If you take a look in the .Rmd file, you can see three R chunks.
In the first chunk called "variables", three random variables were generated.
Each variable has 100 observations.
The variables were created in a way that the variables y and z covary with the variable x, but not with each other.
Change the option of the R chunk "variables" so that it is visible in the output file.


```{r variables, echo=FALSE}
set.seed(123)             # set custom seed for reproducible results
x = rnorm(100)            # generate random variable x with 100 obs.
y = 0.2 * x + rnorm(100)  # generate second random variable y with 100 obs. that covaries with x
z = 0.3 * x + rnorm(100)  # generate third random variable z with 100 obs. that also covaries with x
```

Now calculate the correlations between x and y and x and z in the next chunk.

```{r correlation, echo=FALSE}
# compute correlation between x and y 

# compute correlation between x and z

```

Now, we want to calculate a simple linear regression model.
The formula describing the model is x \~ y + z.
Use the chunk "regression" to run the linear regression.
Report the results in the text body of the document including beta-weights and multiple R-squared.

```{r regression, echo=FALSE}
# compute regression model

```

Finally, change the factors by which y and z are multiplied by x in the chunk "variables".
Knit the document and compare the results in the ew output file with the previous ones.