---
title: 'R Markdown Lesson 01: Getting Started'
subtitle: 'Solutions to the exercises'
author: "Alexander Strobel & Christoph Scheffel"
output: word_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Solutions to the exercises of Lesson 01

## First exercise

Your task was to set up an R Markdown file with Word output and insert a subtitle in the YAML header. The solution to this task can be seen above in the header of this document.

## Second exercise

Here, you should try to write 'partial eta-squared' using LaTeX code. This is done via "\\eta_{p}^2" embedded double Dollar symbols, which is then formatted as $\eta_{p}^2$.

## Third exercise

Your task was to use the variables $x$ and $y$ generated in Lesson 01 to run a simple linear regression in a code chunk and report the results in the body of the document including multiple $R$-squared and the $F$-statistic. The  code chunk below (invisible in the knitted document) shows how this could be done. The resulting texts reads:

```{r cor1, echo = F}
set.seed(242)       # set custom seed for reproducible results
x = rnorm(100)        # generate random normally distributed variable x with 100 observations
y = 0.2 * x + rnorm(100)  # generate random normally distributed variable y with 100 observations that covaries with x

apa.coef <- function(coef, gt1 = T, digits = 2) {
    # function for reporting results according to APA conventions
    if (gt1 == T) {
      coef = format(coef, digits = 2, nsmall = 2)
    } else if (gt1 == F) {
      coef = sub('0.', '.', format(coef, digits = digits, nsmall = digits))
    }
    return(coef)
  }

slm = summary(lm(y ~ x)) # summary of linear regression model
# slope for reporting
slope  = apa.coef(slm$coefficients[2, 1])
# standard error for reporting
SE     = apa.coef(slm$coefficients[2, 2])
# t-value for reporting
tval   = apa.coef(slm$coefficients[2, 3])
# p-value for reporting, either for >= .001 or for < .001
if (slm$coefficients[2, 4] >= .001) {
  pval = paste0("= ", apa.coef(slm$coefficients[2, 4], gt1 = F, digits = 3))
} else {
  pval = "< .001"
}
# classification of p-value
if (slm$coefficients[2, 4] < .05) {
  psig = "significantly"
} else {
  psig = "nonsignificantly"
}
# adjusted R-square for reporting
rsq    = apa.coef(slm$r.squared, gt1 = F, digits = 2)
```

Variable $x$ `r psig` predicted variable $y$ in a linear regression model, $B=$ `r slope`, $SE=$ `r SE`, $t=$ `r tval`, $p$ `r pval`. The explained variance of the model was $R^2=$ `r rsq`.

## Fourth execise

You should write a function that reports the means and standard deviations of $x$ and $y$, either without or with a statement on whether the variables were normally distributed (as determined via a Shapiro-Wilk test, with $p>.20$ indicating no deviation from the normal distribution). This code chunk below (invisible in the knitted document) shows an option. 

```{r descriptives, echo = F}
descriptives <- function(variable, normality.test = T) {
  M = mean(variable)
  SD = sd(variable)
  if (normality.test == T) {
    SW.p = shapiro.test(variable)$p
    if (SW.p > .20) {
      normal = "normally"
    } else {
      normal = "non-normally"
    }
  } else {
    SW.p = NULL
    normal = ""
  }
  return(list(mean = M, sd = SD, normal = normal, p.normal = SW.p
  ))
}

```

You can then use this function to report descriptve statistics: Variable $x$ ($M=$ `r round(descriptives(x)$mean, 2)`, $SD=$ `r round(descriptives(x)$sd, 2)`) was `r descriptives(x)$normal` distributed, Shapiro-Wilk-test, $p=$  `r sub('0.', '.', format(descriptives(x)$p.normal, digits = 3, nsmall = 3))`. For this very purpose (i.e., to give means and standard deviations, you of course do not need a special function), you can simply write: Variable $x$ had a mean of $M=$ `r round(mean(x),2)`. But if you want to have certain text to be inserted to state whether or not some result is significant or not or some distribution is normal or non-normal, such a function makes this an easier job.   

## Fifth exercise

The task was to insert a figure of a scatter plot of the regression of $y$ on $x$ including confidence intervals of the regression slope using the _predict()_ function and also to set the figure caption in the code chunk using the argument fig.cap. Here is, how this could be done.

```{r figure, echo = F, fig.height = 3.5, fig.width = 3.5, fig.align = 'center', fig.cap = "__Figure 1.__ Scatterplot of the variables of interest. "}
plot(x, y, xlim = c(-3, 3), ylim = c(-3, 3), cex = .6, cex.axis = .8, las = 1) # change las to 0 and see what happens to the plot
abline(lm(y ~ x), lwd = 1.5)
text(2.5, -2.5, bquote(italic(r) == .(round(cor(x, y), 2))), cex = .8)
newx = seq(par('usr')[1], par('usr')[2], by = .01)
p = predict(lm(y ~ x), data.frame(x = newx), interval = 'confidence')
lines(newx, p[, 2], lty = 5)
lines(newx, p[, 3], lty = 5)
```

You may notice two things: First, the resolution of the figure is very bad, and it has large white margins. Second, although in the header of the code chunk, it is stated "fig.align='center'", the figure is placed left. Both things have to do with the Word output. Change the output format in the YAML header at the beginning of this document to "pdf_document" and see what happens, when you knit the document again.